---
title: "Forecasting and Predicting Bus Transit Alternatives in El Paso, Texas"
author: "Charlie Huemmler, Yingxue Ou, Jack Rummler"
date: "2023-02-18"
output:
  html_document: 
    toc: true
    toc_float: true
    toc_collapsed: true
    code_folding: hide
    pdf_document: default
    theme: journal
---

```{r setup, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE,
  warning=FALSE,
  message=FALSE,
  results='hide')

library(sf)
library(tidyverse)
library(tidycensus)
library(tigris)
library(boxr)
library(lubridate)
library(here)
library(readxl)
library(gridExtra)
library(scales)
library(viridis)
library(leaflet)
library(ggplotify)
library(ggmap)
library(classInt)
library(rlang)
library(dplyr)
library(maps)
library(plotly)
library(RColorBrewer)
library(htmlwidgets)
library(magick)
library(cowplot)
library(paletteer)
library(stplanr)
library(paletteer)
library(scales)
library(gganimate)
library(transformr)
library(scales)
library(gganimate)
library(glue)
library(ggtext)
library(gapminder)
library(ggplot2)
library(osmdata)
#install.packages("mapview")
library(mapview)

source("https://raw.githubusercontent.com/urbanSpatial/Public-Policy-Analytics-Landing/master/functions.r")

census_api_key("746ea8916547306ae2abf2aafe059e1a1b70b98a", overwrite = TRUE)

data_folder <- file.path(
  here() %>% 
    dirname(), 'data')

mapTheme <- theme(
    text = element_text( color = "black"),
    plot.title = element_text(size = 14,colour = "black"),
    plot.subtitle=element_text(face="italic"),
    plot.caption=element_text(hjust=0),
    axis.ticks = element_blank(),
    panel.background = element_blank(),axis.title = element_blank(),
    axis.text = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(colour = "black", fill=NA, size=2)
  ) 

plotTheme <- theme(
    text = element_text( color = "black"),
    plot.title = element_text(size = 14,colour = "black"),
    plot.subtitle=element_text(face="italic"),
    plot.caption=element_text(hjust=0),
    #axis.ticks = element_blank(),
    panel.background = element_blank(),
    #axis.title = element_blank(),
    #axis.text = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(colour = "black", fill=NA, size=2)
  ) 

palette7 <- c("#B8860B", "#2F4F4F", "#8B0000", "#FFD700", "#556B2F", "#ADD8E6", "#FF69B4")


sf::sf_use_s2(FALSE)
options(scipen=999)
```

# Introduction

This project was created by Charlie Huemmler, Yingxue Ou, and Jack Rummler as part of the University of Pennsylvania Master of Urban Spatial Analytics practicum. We would like to thank our client, Alex Hoffman, AICP, of the El Paso Capital Improvements Department for his continual knowledge, eagerness to help, and enthusiasm for this project. We also would like to thank Professors Matthew Harris and Michael Fichman for their invaluable mentorship throughout this project.

## El Paso Context

Sun Metro is the transportation provider in the city of El Paso, Texas. Sun Metro has made several expansions in the past decade, most notably adding four bus rapid transit (BRT) lines, adding the streetcar network, and a new transit center. With many transit agencies in the last decade, there has been a yearly decrease in bus ridership, with a sharp decrease at the start of the COVID-19 pandemic. Now, Sun Metro experiences about 63% of pre-pandemic ridership numbers, but the agency is looking to explore the implications of new bus transit services.

El Paso is the sixth-largest city in Texas located at the far western tip of the state, just immediately north of Ciudad Juárez. The city is bounded by the Rio Grande River to the south, the Franklin Mountain Range to the north, and Fort Bliss Military Base to the northeast. The context map below shows the geographical constraints where the Franklin Mountains are represented in light green, Fort Bliss Military Base in dark green, and the Rio Grande River in light blue, overlain on top of population density of the city at a census tract level. As we can see, downtown El Paso is notched in between the mountains and river, with heavy sprawl toward the north and east.

[context_map.png]("C:/Users/jtrum/pennmusa/MUSA8010/repository/context_map.png")

The four BRIO lines currently have the greatest ridership and frequency, accounting for 40% of current ridership. Local bus routes and BRIO lines account for 65% of system ridership. BRIO stops are spaced out every 3/4 to 1 mile, thus the agency is seeking ways to connect local bus transit routes to BRIO lines, accommodating issues associated with the first-mile/last-mile problem. Within new bus transit alternatives, Sun Metro wants to maximize both the revenue of routes as well as the equity and accessibility of bus transit options for El Paso residents.

## Use Case

Our client wishes to understand the ramifications of bus transit alternatives on the existing Sun Metro network based on current ridership counts and demographic variables, built environment characteristics, and spatial effects. We are developing a proof-of-concept, unbiased evaluation framework for Sun Metro transit planners to maximize the social equity and financial profitability outcomes of alternative bus route scenarios.

We are building a predictive model of latent bus transit demand. Within our model, we bake in features related to socio-economic spatial characteristics, built environment features, and spatial lags that affect accessibility to transit. By using features related to system accessibility, we bring an equity approach to the model.

# Data Gathering & Analysis

## Current Bus Network

We utilized road centerline data and bus route data obtained from El Paso's Open Data portal to understand the overall transit system's network. The current network is visually represented below, where roads were depicted in black and the bus network was overlain in orange. We can see that the network coverage extended significantly towards areas surrounding downtown El Paso. However, several neighborhoods located in the eastern part of the city remain unserved by the current transit system. This highlights a potential gap in the transit service coverage of the city and underscores the need for equity considerations in ensuring any El Paso resident has access to transit.

```{r current bus network, warning=FALSE, message=FALSE, results=FALSE}
bus_routes <- read_sf(paste(data_folder, "/BusRoutes.geojson", sep = ''))

ggplot()+
  geom_sf(data=road_centerlines, 
          alpha=0.1, 
          color="black", 
          size=1)+
  geom_sf(data=bus_routes,
          alpha=0.9, 
          color="#C96A52FF", 
          size=1)+
  labs(title="SunMetro Bus Network",
       caption="Data: Open Data El Paso")+
  mapTheme
```

## Ridership Data (2022)

Our client provided us with ridership data for the year 2022. Data was collected via sensors on a daily basis, recording number of on-boards and off-boards per stop and route. About 85% of buses were supplied with the sensors, thus the sensors had to be rotated to get the most accurate measures. 

We were also provided Replica data to analyze origin and destination data, aggregated at the census tract level. We gathered data from the American Community Survey and Open Data El Paso to integrate as community demographic and built environment indicators.

```{r cleaning ridership data, warning=FALSE, message=FALSE, results=FALSE}
ridership <- read.csv(paste(data_folder, "/ridership.csv", sep = ''))

ridership$Date <- ridership$Date %>% 
  as.character() %>% 
  substring(2) %>% 
  as.Date(format = "%Y%m%d")

stops <- read.csv(paste(data_folder, "/stops.csv", sep = ''))
stops_sf <- stops %>% 
  st_as_sf(coords = c('stop_lon','stop_lat')) 

riderstops <- read.csv(paste(data_folder, "/riderstops1.csv", sep = ''))

df <- riderstops %>%
  na.omit(riderstops[, c("stop_lat", "stop_lon")]) %>% 
  st_as_sf(coords = c("stop_lon", "stop_lat"), crs = 4269)

df$longitude <- st_coordinates(df$geometry)[, "X"]
df$latitude <- st_coordinates(df$geometry)[, "Y"]

df$RT <- as.numeric(df$RT)

df <- df %>%
  mutate(
    type = case_when(
      RT == 2 ~ "Local",
      RT == 7 ~ "Local",
      RT == 10 ~ "Local",
      RT == 14 ~ "Local",
      RT == 15 ~ "Local",
      RT == 24 ~ "Local",
      RT == 25 ~ "Local",
      RT == 32 ~ "Local",
      RT == 33 ~ "Local",
      RT == 34 ~ "Local",
      RT == 35 ~ "Local",
      RT == 36 ~ "Local",
      RT == 37 ~ "Local",
      RT == 50 ~ "Local",
      RT == 51 ~ "Local",
      RT == 52 ~ "Local",
      RT == 53 ~ "Local",
      RT == 54 ~ "Local",
      RT == 58 ~ "Local",
      RT == 61 ~ "Local",
      RT == 62 ~ "Local",
      RT == 63 ~ "Local",
      RT == 64 ~ "Local",
      RT == 65 ~ "Local",
      RT == 66 ~ "Local",
      RT == 67 ~ "Local",
      RT == 68 ~ "Local",
      RT == 69 ~ "Local",
      RT == 72 ~ "Local",
      RT == 74 ~ "Local",
      RT == 86 ~ "Local",
      RT == 4 ~ "Circulator",
      RT == 8 ~ "Circulator",
      RT == 5 ~ "Express",
      RT == 6 ~ "Express",
      RT == 26 ~ "Express",
      RT == 59 ~ "Express",
      RT == 76 ~ "Express",
      RT == 11 ~ "Feeder",
      RT == 12 ~ "Feeder",
      RT == 13 ~ "Feeder",
      RT == 16 ~ "Feeder",
      RT == 19 ~ "Feeder",
      RT == 43 ~ "Feeder",
      RT == 44 ~ "Feeder",
      RT == 46 ~ "Feeder",
      RT == 56 ~ "Feeder",
      RT == 60 ~ "Feeder",
      RT == 84 ~ "Feeder",
      RT == 89 ~ "Feeder",
      RT == 205 ~ "BRIO",
      RT == 206 ~ "BRIO",
      RT == 207 ~ "BRIO",
      RT == 208 ~ "BRIO",
      RT == 17 ~ "Other",
      RT == 20 ~ "Other",
      RT == 21 ~ "Other",
      RT == 41 ~ "Other",
      RT == 82 ~ "Other",
      RT == 87 ~ "Other",
      RT == 500 ~ "Streetcar"
    )
  )

road_centerlines <- read_sf(paste(data_folder, "/EPCenterline.shp", sep = ''))
```

```{r find stops without ridership data, warning = FALSE, message = FALSE, results=FALSE}
RT_stop_avg <- df %>%
  group_by(RT, TP, longitude, latitude) %>%
  summarize(avg_ons = mean(Ons), 
            avg_offs = mean(Offs))

# RT_stop_avg <- RT_stop_avg %>%
#   mutate(
#     type = case_when(
#       RT == 2 ~ "Local",
#       RT == 7 ~ "Local",
#       RT == 10 ~ "Local",
#       RT == 14 ~ "Local",
#       RT == 15 ~ "Local",
#       RT == 24 ~ "Local",
#       RT == 25 ~ "Local",
#       RT == 32 ~ "Local",
#       RT == 33 ~ "Local",
#       RT == 34 ~ "Local",
#       RT == 35 ~ "Local",
#       RT == 36 ~ "Local",
#       RT == 37 ~ "Local",
#       RT == 50 ~ "Local",
#       RT == 51 ~ "Local",
#       RT == 52 ~ "Local",
#       RT == 53 ~ "Local",
#       RT == 54 ~ "Local",
#       RT == 58 ~ "Local",
#       RT == 61 ~ "Local",
#       RT == 62 ~ "Local",
#       RT == 63 ~ "Local",
#       RT == 64 ~ "Local",
#       RT == 65 ~ "Local",
#       RT == 66 ~ "Local",
#       RT == 67 ~ "Local",
#       RT == 68 ~ "Local",
#       RT == 69 ~ "Local",
#       RT == 72 ~ "Local",
#       RT == 74 ~ "Local",
#       RT == 86 ~ "Local",
#       RT == 4 ~ "Circulator",
#       RT == 8 ~ "Circulator",
#       RT == 5 ~ "Express",
#       RT == 6 ~ "Express",
#       RT == 26 ~ "Express",
#       RT == 59 ~ "Express",
#       RT == 76 ~ "Express",
#       RT == 11 ~ "Feeder",
#       RT == 12 ~ "Feeder",
#       RT == 13 ~ "Feeder",
#       RT == 16 ~ "Feeder",
#       RT == 19 ~ "Feeder",
#       RT == 43 ~ "Feeder",
#       RT == 44 ~ "Feeder",
#       RT == 46 ~ "Feeder",
#       RT == 56 ~ "Feeder",
#       RT == 60 ~ "Feeder",
#       RT == 84 ~ "Feeder",
#       RT == 89 ~ "Feeder",
#       RT == 205 ~ "BRIO",
#       RT == 206 ~ "BRIO",
#       RT == 207 ~ "BRIO",
#       RT == 208 ~ "BRIO",
#       RT == 17 ~ "Other",
#       RT == 20 ~ "Other",
#       RT == 21 ~ "Other",
#       RT == 41 ~ "Other",
#       RT == 82 ~ "Other",
#       RT == 87 ~ "Other",
#       RT == 500 ~ "Streetcar"
#     )
#   )
```

We can see from the bar chart that BRIO routes are predominantly the most popular type of transit. Because of the fixed nature of BRIO routes, and political implications of bus transit design, we will primarily focus on local routes as part of our analysis. While some local routes experience relatively high ridership, we see total ridership peter out, with opportunity to increase ridership to underperforming routes.

```{r route summary, warning = FALSE, message = FALSE, results = FALSE}
RT_avg_sum <- df %>%
  group_by(RT) %>%
  summarize(avg_ons = mean(Ons), 
             avg_offs = mean(Offs),
             total_ons = sum(Ons),
             total_offs = sum(Offs)) %>%
  mutate(total_ridership = total_ons + total_offs) %>%
  mutate(
    type = case_when(
      RT == 2 ~ "Local",
      RT == 7 ~ "Local",
      RT == 10 ~ "Local",
      RT == 14 ~ "Local",
      RT == 15 ~ "Local",
      RT == 24 ~ "Local",
      RT == 25 ~ "Local",
      RT == 32 ~ "Local",
      RT == 33 ~ "Local",
      RT == 34 ~ "Local",
      RT == 35 ~ "Local",
      RT == 36 ~ "Local",
      RT == 37 ~ "Local",
      RT == 50 ~ "Local",
      RT == 51 ~ "Local",
      RT == 52 ~ "Local",
      RT == 53 ~ "Local",
      RT == 54 ~ "Local",
      RT == 58 ~ "Local",
      RT == 61 ~ "Local",
      RT == 62 ~ "Local",
      RT == 63 ~ "Local",
      RT == 64 ~ "Local",
      RT == 65 ~ "Local",
      RT == 66 ~ "Local",
      RT == 67 ~ "Local",
      RT == 68 ~ "Local",
      RT == 69 ~ "Local",
      RT == 72 ~ "Local",
      RT == 74 ~ "Local",
      RT == 86 ~ "Local",
      RT == 4 ~ "Circulator",
      RT == 8 ~ "Circulator",
      RT == 5 ~ "Express",
      RT == 6 ~ "Express",
      RT == 26 ~ "Express",
      RT == 59 ~ "Express",
      RT == 76 ~ "Express",
      RT == 11 ~ "Feeder",
      RT == 12 ~ "Feeder",
      RT == 13 ~ "Feeder",
      RT == 16 ~ "Feeder",
      RT == 19 ~ "Feeder",
      RT == 43 ~ "Feeder",
      RT == 44 ~ "Feeder",
      RT == 46 ~ "Feeder",
      RT == 56 ~ "Feeder",
      RT == 60 ~ "Feeder",
      RT == 84 ~ "Feeder",
      RT == 89 ~ "Feeder",
      RT == 205 ~ "BRIO",
      RT == 206 ~ "BRIO",
      RT == 207 ~ "BRIO",
      RT == 208 ~ "BRIO",
      RT == 17 ~ "Other",
      RT == 20 ~ "Other",
      RT == 21 ~ "Other",
      RT == 41 ~ "Other",
      RT == 82 ~ "Other",
      RT == 87 ~ "Other",
      RT == 500 ~ "Streetcar"
    )
  )
RT_avg_sum_st <- st_drop_geometry(RT_avg_sum)

transit_lines <- st_read(paste(data_folder, "/transit_lines.geojson", sep = ''))
merged <- merge(transit_lines, RT_avg_sum_st, by.x = "route_short_name", by.y = "RT")

RT_avg_sum_st <- RT_avg_sum_st[order(-RT_avg_sum_st$total_ridership),]

ggplot(RT_avg_sum_st, aes(y = total_ridership, x = reorder(factor(RT), -total_ridership), fill = type)) +
  geom_bar(stat = "identity") +
  xlab("Total Ridership") +
  ylab("Route") +
  scale_fill_manual(values = palette7, name="Type of Bus", expand=c(0.2, 0)) +
  labs(title="Total Ridership by Route and Type",
       caption="Data: El Paso Capital Improvements Department, 2022")+
  plotTheme+
  theme(axis.text.x = element_text(angle=90))
```

As the scatter plot indicates, there is a strong correlation between average on-board and off-board frequency. This indicates that in bus transit trips, the average rider may exhibit round-trip behavior, boarding the bus at point A to travel to point B and then returning to point A without any intermediate stops.

```{r scatterplot1, warning=FALSE, message=FALSE, results=FALSE}
RT_stop_avg_st <- st_drop_geometry(RT_stop_avg)

ggplot(RT_stop_avg_st, aes(x = avg_ons, y = avg_offs)) +
  geom_point() +
  labs(x = "Average on-boarding", 
       y = "Average off-boarding", 
       title = "Scatterplot of on-boarding and off-boarding", 
       subtitle="Average values grouped by route and stop information",
       caption="Data: El Paso Capital Improvements Department, 2022")+
  geom_smooth(method = lm, se=FALSE, colour = "#C96A52FF", size=1, )+
  geom_abline(intercept=0, slope=1, color="#C96A52FF", alpha=0.3, size=1, style='dashed')+
  plotTheme
```


```{r total offboards per route, warning = FALSE, message = FALSE, results=FALSE}
ggplot()+
  geom_sf(data=road_centerlines, 
          alpha=0.1, 
          color="black", 
          size=1)+
  geom_sf(data=merged,
          aes(color=total_offs),
          size=1)+
  scale_color_paletteer_c("grDevices::Red-Yellow", 
                          -1,
                          name="Sum")+
  labs(title="Total Offboards per route (2022)")+
  mapTheme
```

Highest onboarding and offboarding ridership is among BRIO lines. There seems to be a spatial pattern where lines that go further out into the suburbs experience less overall ridership.

## Leaflet Maps

The leaflet maps below are interactive to understand spatial distribution of current ridership patterns at a route and stop level.

```{r leaflet 1, warning = FALSE, message = FALSE, results=FALSE}
jenks_breaks <- classIntervals(RT_stop_avg$avg_ons, n = 8, style = "jenks")$brks

colors <- c("#7D0112FF", "#9A4221FF", "#BF7B42FF", "#E0BA79FF", "#F0E4B0FF", "#C0C0C0FF", "#787878FF", "#4B4B4BFF", "#000000FF")

# LL1 <- leaflet() %>%
#   addTiles() %>%
#   setView(lng = -106.485, lat = 31.763, zoom = 12) %>%
#   addCircleMarkers(data = RT_stop_avg, lng = ~longitude, lat = ~latitude, 
#                    radius = 5, 
#                    color = ~colors[cut(avg_ons, breaks = jenks_breaks)], 
#                    fillOpacity = 0.7,
#                    popup = paste(
#                      "Route: ", RT_stop_avg$RT, "<br>",
#                      "Stop: ", RT_stop_avg$TP, "<br>",
#                      "Average Onboarding: ", round(RT_stop_avg$avg_ons, digits = 2))) %>%
#   addLegend("bottomright", 
#             title = "Average Onboarding per stop & route", 
#             colors = colors, 
#             labels = c(sprintf("%.2f - %.2f", jenks_breaks[1], jenks_breaks[2]), 
#                        sprintf("%.2f - %.2f", jenks_breaks[2], jenks_breaks[3]), 
#                        sprintf("%.2f - %.2f", jenks_breaks[3], jenks_breaks[4]), 
#                        sprintf("%.2f - %.2f", jenks_breaks[4], jenks_breaks[5]), 
#                        sprintf("%.2f - %.2f", jenks_breaks[5], jenks_breaks[6]), 
#                        sprintf("%.2f - %.2f", jenks_breaks[6], jenks_breaks[7]), 
#                        sprintf("%.2f - %.2f", jenks_breaks[7], jenks_breaks[8]), 
#                        sprintf("%.2f - %.2f", jenks_breaks[8], jenks_breaks[9]), 
#                        sprintf(">%0.2f", jenks_breaks[9])),
#             opacity = 1) 
# 
# mapview(map=LL1, 
#         zcol="avg_ons", 
#         fill = c("#7D0112FF", "#9A4221FF", "#BF7B42FF", "#E0BA79FF", "#F0E4B0FF", "#C0C0C0FF", "#787878FF", "#4B4B4BFF", "#000000FF"), 
#         popup= list(Route = RT_stop_avg$RT,
#                     Stop = RT_stop_avg$TP,
#                     Average_Onboarding = round(RT_stop_avg$avg_ons, digits = 2)), 
#         layer.name = "Average Onboarding")

mapview(RT_stop_avg, 
        zcol = "avg_ons",
        col.regions = c("#7D0112FF", "#9A4221FF", "#BF7B42FF", "#E0BA79FF", "#F0E4B0FF"),
        layer.name = "Average Onboarding", 
        popup = list(Route = RT_stop_avg$RT, Stop = RT_stop_avg$TP, Avg_Ons = round(RT_stop_avg$avg_ons, digits = 2)))
```

```{r leaflet2, warning = FALSE, message = FALSE, results=FALSE}
jenks_breaksOffs <- classIntervals(dfRTx$avg_offs, n = 9, style = "jenks")$brks

LL2 <- leaflet(dfRTx) %>%
  addTiles() %>%
  setView(lng = -106.485, lat = 31.763, zoom = 12) %>%
  addCircleMarkers(data = dfRTx, lng = ~longitude, lat = ~latitude, 
                   radius = 5, 
                   color = ~colors(9)[cut(avg_offs, breaks = jenks_breaksOffs)], 
                   fillOpacity = 0.7,
                   popup = paste(
                     "Route: ", dfRTx$RT, "<br>",
                     "Stop: ", dfRTx$stop_desc, "<br>",
                     "Average Offboarding: ", round(dfRTx$avg_offs, digits = 2))) %>%
  addLegend("bottomright", 
            title = "Average Offboarding per stop & route", 
            colors = colors(9), 
            labels = c(sprintf("%.2f - %.2f", jenks_breaksOffs[1], jenks_breaksOffs[2]), 
                       sprintf("%.2f - %.2f", jenks_breaksOffs[2], jenks_breaksOffs[3]), 
                       sprintf("%.2f - %.2f", jenks_breaksOffs[3], jenks_breaksOffs[4]), 
                       sprintf("%.2f - %.2f", jenks_breaksOffs[4], jenks_breaksOffs[5]), 
                       sprintf("%.2f - %.2f", jenks_breaksOffs[5], jenks_breaksOffs[6]), 
                       sprintf("%.2f - %.2f", jenks_breaksOffs[6], jenks_breaksOffs[7]), 
                       sprintf("%.2f - %.2f", jenks_breaksOffs[7], jenks_breaksOffs[8]), 
                       sprintf("%.2f - %.2f", jenks_breaksOffs[8], jenks_breaksOffs[9]), 
                       sprintf(">%0.2f", jenks_breaksOffs[9])),
            opacity = 1) 
LL2
```

[LL2]("C:/Users/jtrum/pennmusa/MUSA8010/repository/ElPaso-Bus-Network/LL1.html")

# Feature Engineering

## Census Data (American Community Survey 2020)

Next, we analyzed census data at the tract level from the 2020 American Community Survey. We particularly focused on demographic and socioeconomic indicators that we predicted to be correlated to transit ridership demand.

```{r census vars, warning=FALSE, message=FALSE, results=FALSE}
acs2020vars <- load_variables(2020, "acs5", cache = TRUE)

censusvarsEP <- c(
  "B01001_001E", # ACS total Pop estimate
  "B01001I_001E", # Population - Hispanic or Latino 
  "B02001_002E", # Population - White alone
  "B02001_003E", # Population - Black or African American alone
  "B02001_004E", # Population - American Indian and Alaska Native alone
  "B02001_005E", # Population - Asian alone
  "B02001_006E", # Population - Native Hawaiian and Other Pacific Islander alone
  "B02001_007E", # Population - Some other race alone
  "B02001_008E", # Population - Two or more races
  "B19013_001E", # Median household income in the past 12 months (in 2019 inflation-adjusted dollars)
  "B25001_001E", # Total housing units
  "B25002_002E", # Occupancy status - Occupied housing units
  "B25024_001E", # Gross rent as a percentage of household income in the past 12 months
  "B25044_001E", # Vehicles available
  "B28005_001E", # Means of transportation to work by age
  "B28010_001E", # Commuting time to work (in minutes)
  "B10052_002E", # Disability
  "B06009_002E", # Less than high school
  "B06009_003E", # High School
  "B06009_004E", # Associates or equivalent
  "B06009_005E", # Bachelors
  "B06009_006E" # Graduate or prof. degree
)

elpaso <- get_acs(geography = "tract",
                             year = 2020, 
                             variables = censusvarsEP, 
                             geometry = T,
                             state = "TX", 
                             county = "El Paso", 
                             output = "wide") 

elpaso <- elpaso %>%
rename(
  totalPop = B01001_001E, 
  hlPop = B01001I_001E, 
  whitePop = B02001_002E, 
  blackPop = B02001_003E,
  aiPop = B02001_004E,
  asianPop = B02001_005E,
  nhPop = B02001_006E,
  otherRacePop = B02001_007E,
  twoPlusRacePop = B02001_008E,
  medHHInc = B19013_001E,
  totalHU = B25001_001E, 
  occupiedHU = B25002_002E, 
  grossRentPerInc = B25024_001E, 
  transByAge = B28005_001E,
  commuteToWork = B28010_001E,
  lessThanHS = B06009_002E,
  highSchool = B06009_003E,
  associatesDeg = B06009_004E,
  bachelorDeg = B06009_005E,
  professionalDeg = B06009_006E,
  disability = B10052_002E)
  
mutate(
  area_sqmile = st_area(geometry)/2590000,
  vacancyPct = total_vacant.2020/area_sqmile,
      pctWhite.2020 = total_White.2020/area_sqmile,
      pctHL.2020 = total_HL.2020/area_sqmile,
      pctNV.2020 = noV.2020/total_HU.2020*100,
      pctUnder18.2020 = under18.2020/total_pop.2020,
      pctDisAb.2020 = disAb.2020/total_pop.2020,
      pop_dense = as.numeric(total_pop.2020 / area_sqmile)
  ) %>%
  st_as_sf(crs = 4269) 

```


## Median household income
```{r medhhinc, warning = FALSE, cache = TRUE}

medHHInc <- 
  ggplot()+
  geom_sf(data = elpaso, aes(fill =  med_HH_Income.2020),
          color = "transparent")+
  geom_sf(data = elpaso %>%
          st_union(),
          color = "white",
          fill = "transparent")+
  scale_fill_paletteer_c("grDevices::Red-Yellow", -1,
                          name="$")+
  geom_point(data = dfRTx_5on, aes(x = longitude, y = latitude, color = "avg_ons"), alpha=0.4) +
  geom_point(data = dfRTx_5off, aes(x = longitude, y = latitude, color = "avg_offs"), alpha=0.4) +
  scale_color_manual(name = "Stop Ridership Data", 
                     values = c("avg_ons" = "limegreen", "avg_offs" = "skyblue"),
                     labels = c("Average Ons > 5", "Average Offs > 5")) +
  labs(
    title = "Median household income",
    subtitle = "Data: US Census Bureau, ACS 5-year estimates (2020)")+
  mapTheme

medHHInc
```

## % White Population

```{r plot_noV1, warning = FALSE, cache = TRUE}
pctWhite <- 
  ggplot()+
  geom_sf(data = elpaso, aes(fill =  pctWhite.2020),
          color = "transparent")+
  geom_sf(data = elpaso %>%
          st_union(),
          color = "white",
          fill = "transparent")+
  scale_fill_paletteer_c("grDevices::Red-Yellow", -1,
                          name="%")+
  geom_point(data = dfRTx_5on, aes(x = longitude, y = latitude, color = "avg_ons"), alpha=0.4) +
  geom_point(data = dfRTx_5off, aes(x = longitude, y = latitude, color = "avg_offs"), alpha=0.4) +
  scale_color_manual(name = "Stop Ridership Data", 
                     values = c("avg_ons" = "limegreen", "avg_offs" = "skyblue"),
                     labels = c("Average Ons > 5", "Average Offs > 5")) +
  labs(
    title = "Identifies race as white",
    subtitle = "Data: US Census Bureau, ACS 5-year estimates (2020)")+
  mapTheme

pctWhite
```

## % Vacant Households

```{r}
pctVacant <- ggplot()+
  geom_sf(data = elpaso, aes(fill = vacancyPct.2020),
          color = "transparent")+
  geom_sf(data = elpaso %>%
          st_union(),
          color = "white",
          fill = "transparent")+
  scale_fill_paletteer_c("grDevices::Red-Yellow", -1,
                          name="%")+
  geom_point(data = dfRTx_5on, aes(x = longitude, y = latitude, color = "avg_ons"), alpha=0.4) +
  geom_point(data = dfRTx_5off, aes(x = longitude, y = latitude, color = "avg_offs"), alpha=0.4) +
  scale_color_manual(name = "Stop Ridership Data", 
                     values = c("avg_ons" = "limegreen", "avg_offs" = "skyblue"),
                     labels = c("Average Ons > 5", "Average Offs > 5")) +
  labs(
    title = "Vacant Households",
    subtitle = "Data: US Census Bureau, ACS 5-year estimates (2020)")+
  mapTheme

pctVacant
```

## % No Vehicle Households

```{r pct no vehicle}
pctNoVehicle <- ggplot()+
  geom_sf(data = elpaso, aes(fill = pctNV.2020),
          color = "transparent")+
  geom_sf(data = elpaso %>%
          st_union(),
          color = "white",
          fill = "transparent")+
  scale_fill_paletteer_c("grDevices::Red-Yellow", -1,
                          name="%")+
  geom_point(data = dfRTx_5on, aes(x = longitude, y = latitude, color = "avg_ons"), alpha=0.4) +
  geom_point(data = dfRTx_5off, aes(x = longitude, y = latitude, color = "avg_offs"), alpha=0.4) +
  scale_color_manual(name = "Stop Ridership Data", 
                     values = c("avg_ons" = "limegreen", "avg_offs" = "skyblue"),
                     labels = c("Average Ons > 5", "Average Offs > 5")) +
  labs(
    title = "Zero-vehicle households",
    subtitle = "Data: US Census Bureau, ACS 5-year estimates (2020)")+
  mapTheme

pctNoVehicle
```

## % Hispanic/Latino Population

```{r hispanic latino pop}
pctHL <- ggplot()+
  geom_sf(data = elpaso, aes(fill = pctHL.2020),
          color = "transparent")+
  geom_sf(data = elpaso %>%
          st_union(),
          color = "white",
          fill = "transparent")+
  scale_fill_paletteer_c("grDevices::Red-Yellow", -1,
                          name="%")+
  geom_point(data = dfRTx_5on, aes(x = longitude, y = latitude, color = "avg_ons"), alpha=0.4) +
  geom_point(data = dfRTx_5off, aes(x = longitude, y = latitude, color = "avg_offs"), alpha=0.4) +
  scale_color_manual(name = "Stop Ridership Data", 
                     values = c("avg_ons" = "limegreen", "avg_offs" = "skyblue"),
                     labels = c("Average Ons > 5", "Average Offs > 5")) +
  labs(
    title = "Hispanic/Latino Population",
    subtitle = "Data: US Census Bureau, ACS 5-year estimates (2020)")+
  mapTheme

pctHL
```

## % Residents under the age of 18

```{r pct under 18}
pctUnder18 <- ggplot()+
  geom_sf(data = elpaso, aes(fill = pctUnder18.2020),
          color = "transparent")+
  geom_sf(data = elpaso %>%
          st_union(),
          color = "white",
          fill = "transparent")+
  scale_fill_paletteer_c("grDevices::Red-Yellow", -1,
                          name="%")+
  geom_point(data = dfRTx_5on, aes(x = longitude, y = latitude, color = "avg_ons"), alpha=0.4) +
  geom_point(data = dfRTx_5off, aes(x = longitude, y = latitude, color = "avg_offs"), alpha=0.4) +
  scale_color_manual(name = "Stop Ridership Data", 
                     values = c("avg_ons" = "limegreen", "avg_offs" = "skyblue"),
                     labels = c("Average Ons > 5", "Average Offs > 5")) +
  labs(
    title = "% of residents under the age of 18",
    subtitle = "Data: US Census Bureau, ACS 5-year estimates (2020)")+
  mapTheme

pctUnder18
```

## % Disabled population

```{r pct disab}
pctDisabled <- ggplot()+
  geom_sf(data = elpaso, aes(fill = pctDisAb.2020),
          color = "transparent")+
  geom_sf(data = elpaso %>%
          st_union(),
          color = "white",
          fill = "transparent")+
  scale_fill_paletteer_c("grDevices::Red-Yellow", -1,
                          name="%")+
  geom_point(data = dfRTx_5on, aes(x = longitude, y = latitude, color = "avg_ons"), alpha=0.4) +
  geom_point(data = dfRTx_5off, aes(x = longitude, y = latitude, color = "avg_offs"), alpha=0.4) +
  scale_color_manual(name = "Stop Ridership Data", 
                     values = c("avg_ons" = "limegreen", "avg_offs" = "skyblue"),
                     labels = c("Average Ons > 5", "Average Offs > 5")) +
  labs(
    title = "% of residents with a disability",
    subtitle = "Data: US Census Bureau, ACS 5-year estimates (2020)")+
  mapTheme

pctDisabled
```



```{r open data el paso loading}
road_centerlines <- read_sf(paste(data_folder, "/EPCenterline.shp", sep = ''))

bike_lanes <- read_sf(paste(data_folder, "/BikeLanesFinal.shp", sep = ''))
bike_lanesEx <- bike_lanes[bike_lanes$Status != "PROPOSED", ]
bike_lanesPr <- bike_lanes[bike_lanes$Status != "EXISTING", ]

parks <- read_sf(paste(data_folder, "/ParkandRecreationFacilities.shp", sep = ''))

#schools <- read_sf(paste(data_folder, "/Schools.shp", sep = ''))

#futureLU <- read_sf(paste(data_folder, "/FutureLandUse.shp", sep = ''))

TIRZ <- read_sf(paste(data_folder, "/TIRZ.shp", sep = ''))
```

## Existing Bike Lanes + Other Roadway Infrastructure

```{r bike lanes}
ggplot()+
  geom_sf(data=road_centerlines, alpha=0.1)+
  geom_sf(data=bike_lanesEx, aes(color="Status"), size=1, color="#8E063BFF")+
  geom_sf(data=bike_lanesPr, aes(color="Status"), size=1, color="#E5E797FF", alpha=0.4)+
  geom_point(data = dfRTx_5on, aes(x = longitude, y = latitude, color = "avg_ons"), alpha=0.1) +
  geom_point(data = dfRTx_5off, aes(x = longitude, y = latitude, color = "avg_offs"), alpha=0.1) +
  scale_color_manual(name = "Stop Ridership Data", 
                     values = c("avg_ons" = "limegreen", "avg_offs" = "skyblue"),
                     labels = c("Average Ons > 5", "Average Offs > 5")) +
  labs(title="Bike Lanes in El Paso, Texas",
       subtitle="Data: Open Data El Paso",
       caption="Lines in red indicate existing bike lanes\nLines in yellow indicate proposed bike lanes")+
  theme()+
  mapTheme
```

## Parks and Recreation Facilities

```{r parks}
parks$Type <- "park"

ggplot()+
  geom_sf(data=road_centerlines, alpha=0.1)+
  geom_sf(data=parks, 
          aes(fill=Type), 
          size=1, 
          alpha=.7, 
          color="#C96A52FF")+
  geom_point(data = dfRTx_5on, aes(x = longitude, y = latitude, color = "avg_ons"), alpha=0.1) +
  geom_point(data = dfRTx_5off, aes(x = longitude, y = latitude, color = "avg_offs"), alpha=0.1) +
  scale_color_manual(name = "Stop Ridership Data", 
                     values = c("avg_ons" = "limegreen", "avg_offs" = "skyblue"),
                     labels = c("Average Ons > 5", "Average Offs > 5")) +
  labs(title="Parks in El Paso, Texas",
       subtitle="Data: Open Data El Paso")+
  mapTheme
```

## Tax Increment Reinvestment Zones

```{r}
TIRZ$TIRZ = "TIRZ"
ggplot()+
  geom_sf(data=road_centerlines, alpha=0.1)+
  geom_sf(data=TIRZ, aes(fill="TIRZ"), size=1, color="#C96A52FF", alpha=.7)+
  geom_point(data = dfRTx_5on, aes(x = longitude, y = latitude, color = "avg_ons"), alpha=0.1) +
  geom_point(data = dfRTx_5off, aes(x = longitude, y = latitude, color = "avg_offs"), alpha=0.1) +
  scale_color_manual(name = "Stop Ridership Data", 
                     values = c("avg_ons" = "limegreen", "avg_offs" = "skyblue"),
                     labels = c("Average Ons > 5", "Average Offs > 5")) +
  labs(title="Tax Increment Reinvestment Zones",
       subtitle="Data: Open Data El Paso")+
  mapTheme
```



## National Walkability Index (2019)

```{r nwi data, }
blocks <- blocks(state = "TX", county = "El Paso") %>%
  st_geometry(blocks)

nwi <- read_csv(paste(data_folder, "/nwi2019.csv", sep = '')) %>%
  dplyr::filter(CBSA_Name == "El Paso, TX") %>%
  dplyr::select(c("OBJECTID",
                 "GEOID10",
                 "TRACTCE",
                 "BLKGRPCE",
                 "TotPop",
                 "CountHU", # number of housing units
                 "P_WrkAge", # percent working age pop. (18-64)
                 "HH", # households
                 "Pct_AO0", # % of HH owning 0 vehicles
                 "Pct_AO1", # % of HH owning 1 vehicle
                 "Pct_AO2p", # % of HH owning 2+ vehicles
                 "R_PCTLOWWAGE", # % of low wage workers
                 "D1C", # employment density
                 "D2R_JOBPOP", # jobs to population balance
                 "D2B_E8MIXA", # employment mix
                 "D2A_EPHHM", # employment + household mix
                 "D3B", # intersection density
                 "D4A", # distance from population-weighted centroid to nearest transit stop
                 "D5AR", # regional accessibility for access to jobs
                 "D5BR", # destination accessibility via transit
                 )) %>%

nwi$GEOID <- paste0(substr(nwi$GEOID10, 1, 5), "0", sprintf("%05d", nwi$TRACTCE), nwi$BLKGRPCE)

nwi_bg <- left_join(nwi, bg, by = "GEOID")

nwi_bg <- st_as_sf(nwi_bg)

nwi_bg %>%
  rename(
    totalPop = TotPop,
    housingUnits = CountHU,
    pctWorkingAge = P_WrkAge,
    housingUnits = CountHU,
    pct0Car = Pct_AO0,
    pct1Car = Pct_AO1,
    pct2Car = Pct_AO2p,
    pctLowWage = R_PCTLOWWAGE,
    employmentDensity = D1C,
    jobsPopBalance = D2R_JOBPOP,
    employmentMix = D2B_E8MIXA,
    employmentHHMix = D2A_EPHHM,
    intersectionDensity = D3B,
    nearestTransit = D4A,
    jobAccessibility = D5AR,
    destinationAccessibility = D5BR
  ) %>%
  dplyr::select(-c("OBJECTID",
                 "GEOID10",
                 "TRACTCE",
                 "BLKGRPCE",))

ggplot(data = nwi_bg) +
  geom_sf(aes(fill = Pct_AO0)) +
  mapTheme


```

```{r ndi api}
# Load the required packages
library(tidycensus)
library(httr)

# Set the Walk Score API key
walkscore_key <- "97c51210c4c9acfce3869b13a5af3f28"

# Get the addresses for all census block groups in El Paso, TX
bg <- get_acs(geography = "block group", 
              variables = "B01003_001", 
              year = 2019,
              state = "TX", 
              county = "El Paso", geometry=TRUE)

nwi19 <- rbind(nwi, bg)
```


## Jobs Data

## Open Street Map

```{r osm amenities}
bbox <- c(left = -106.6435, bottom = 31.6815, right = -106.1845, top = 31.9865)

el_paso_amenities <- osmdata::osmdata_sf(
  bbox = bbox,
  type = "amenity"
)

# Print the first few rows of the resulting data
head(el_paso_amenities$osm_points)

```


# Modeling

# Bus Network and Maps


```{r}
# 
# g <- ggplot()+
#   geom_sf(data = briolines %>% select(-route_long_name), size = 1.5, aes(color = route_id))+
# 
#   geom_sf(data = notbrio %>% select(-route_long_name), alpha = .75, color = 'grey')+
#   geom_sf(data = notbrio, aes(group = route_long_name), color = 'red')+
#   mapTheme+
#   scale_color_manual(values = c('#B0C72C','#660066','#00A8E7'))+
#   labs(title = 'Sun Metro Local Bus Network',
#        subtitle = '{closest_state}')+
#   theme(legend.position = 'none')+
#   transition_states(route_long_name)+
#   exit_fade()
# 
# animate(g, duration = 15, fps = 10, bg = 'black')
# 
# anim_save('citybusses.gif', g, path = "C:/Users/cchue/Documents/Penn MUSA/Spring/Practicum",  duration = 15, fps = 10, bg = 'black')
```







# Modeling

We are building a zero-inflated Poisson regression model, aggregating our ridership data and engineered features into hexbins, to predict latent demand for bus services.

```{r}

elpaso_outline <- elpaso %>% st_union() %>% sf::st_cast()

hex <- st_make_grid(elpaso_outline, cellsize = .01, crs = 4269,  square = F)  %>%  st_sf() 

hex <- hex[elpaso,] %>%
  mutate(uniqueID = rownames(.))
```

```{r}

riderstops_sf <- read_csv(paste(data_folder, "/riderstops1.csv", sep = '')) %>%
  filter(!is.na(stop_lat)) %>%
  st_as_sf(coords = c('stop_lon', 'stop_lat'), crs = 4269) 

stop_riders_agg <- riderstops_sf %>% group_by(TP) %>% summarise(ridership = sum(Ons) + sum(Offs))

ridership_net <- stop_riders_agg %>% 
  dplyr::select(ridership) %>% 
  aggregate(., hex, sum) %>%
  mutate(ridership = replace_na(ridership, 0))

```

```{r}
ggplot(ridership_net %>% st_crop(y= bounds) %>% filter(ridership > 0))+
  geom_sf(aes(fill = ridership), color = NA)+
  geom_sf(data = ridership_net %>% st_crop(y= bounds) %>%  filter(ridership == 0), color = NA, fill = '#fbfae4')+
  scale_fill_paletteer_c("grDevices::Red-Yellow", -1, labels = comma)+
  labs(title = "Ridership Per Hexbin", subtitle = "Downtown has an exceptional amount of ridership")+
  mapTheme
```
